---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Load the data:

```{r}
echo = TRUE
setwd("C:/Users/Kelvin Lin/datasciencecoursera")
activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Sum total nuber of steps for every single date:

```{r, echo=TRUE}
total <- tapply(activity$steps, activity$date, sum, na.rm=T)
```

Plot histogram of the total number of steps taken each day

```{r, echo=TRUE}
hist(total, xlab = "sum of steps per day", main = "Histogram of no. of steps per day" )
```

Calculate mean and median of the total number of steps taken each day:

```{r, echo=TRUE}
mean_total <- round(mean(total))
print(c("The mean is", mean_total))
median_total <- round(median(total))
print(c("The median is", median_total))
```

## What is the average daily activity pattern?
A time series plot of the 5-minute interval and the average number of steps taken is show below:

```{r, echo=TRUE}
min_int <- tapply(activity$steps, activity$interval, mean, na.rm=T)
plot(min_int ~ unique(activity$interval), type="l", xlab = "5-min interval")
```

The 5 min interval that contains the maximum number of steps is:

```{r, echo=TRUE}
min_int[which.max(min_int)]
```

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

To visualise where the number of NAs are:

```{r, echo=TRUE}
table(is.na(activity) == TRUE)
summary(activity)
```

All of the 2304 NAs are in the steps variable.

## Strategy to fill in all of the missing values in the dataset

The following strategy is chosen: for any NA value in the step variable, the mean ( number of steps) of the corresponding interval is taken as the replacing value.

The 'min_int' contains the mean for each single interval calculated over the 61 days. The right value coming from 'min_int' is going to be used to replace the NA at the same interval.

Create a new dataset that will have the NA values replaced by the mean.

```{r, echo=TRUE}
activity2 <- activity  # creation of the dataset that will have no more NAs
for (i in 1:nrow(activity)){
    if(is.na(activity$steps[i])){
        activity2$steps[i]<- min_int[[as.character(activity[i, "interval"])]]
    }
}
```

The histogram of the total number of steps taken each day (for new datase) is shown below:

```{r, echo=TRUE}
total2 <- tapply(activity2$steps, activity2$date, sum, na.rm=T)
hist(total2, xlab = "sum of steps per day", main = "histogram of steps per day")
```

Mean and median values of the new dataset are shown below

```{r, echo=TRUE}
mean_total2 <- round(mean(total2))
print(c("The mean is", mean_total2))
median_total2 <- round(median(total2))
print(c("The median is", median_total2))
```

In order to compare the mean and median values of the old and new dataset:

```{r, echo=TRUE}
df_summary <- rbind(data.frame(mean = c(mean_total, mean_total2), median = c(median_total, median_total2)))
rownames(df_summary) <- c("Old (with NAs)", "New (w/o NAs)")
df_summary
```

Impact of inputing the missing data on the estimates of the total daily number of steps

```{r, echo=TRUE}
summary(activity2)
summary(activity)
```

There are no more NAs.


## Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

Add 2 new columns to the data frame, col weekday_1 converting the date into the correspoding day of the week and col weekday containing the factor "weekend" or "weekday"

```{r, echo=TRUE}
activity2$date <- as.Date(activity2$date, format= "%Y-%m-%d")
activity2$weekday_1 <- weekdays(activity2$date)
for (i in 1:nrow(activity2)){
    if(activity2$weekday_1[i] == "Sunday" | activity2$weekday_1[i] == "Saturday"){
        activity2$weekday[i]<- "weekend"
    } else {

    activity2$weekday[i]<- "weekday"
    }
}
```

To remove col activity2$weekday_1

```{r, echo=TRUE}
activity2$weekday_1 <- NULL
```

### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

Subset data for weekends and weekends, and aggregate data based on mean steps taken for each interval

```{r, echo=TRUE}
activity2_weekend <- subset(activity2, activity2$weekday == "weekend")
activity2_weekday <- subset(activity2, activity2$weekday == "weekday")

mean_activity2_weekday <- tapply(activity2_weekday$steps, activity2_weekday$interval, mean)
mean_activity2_weekend <- tapply(activity2_weekend$steps, activity2_weekend$interval, mean)
```

Use Lattice to plot graph

```{r, echo=TRUE}
library(lattice)
df_weekday <- NULL
df_weekend <- NULL
df_final <- NULL
df_weekday <- data.frame(interval = unique(activity2_weekday$interval), avg = as.numeric(mean_activity2_weekday), day = rep("weekday", length(mean_activity2_weekday)))
df_weekend <- data.frame(interval = unique(activity2_weekend$interval), avg = as.numeric(mean_activity2_weekend), day = rep("weekend", length(mean_activity2_weekend)))
df_final <- rbind(df_weekday, df_weekend)
xyplot(avg ~ interval | day, data = df_final, layout = c(1, 2), type = "l", ylab = "Number of Steps")
```
